---
title: "H9 RPE YOUNG Sample Analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
# R Environment Setup
This analysis uses Seurat v3 and its associated workflows.

```{r}
# All packages are available via CRAN
# Single cell analysis
library(Seurat)

# Plotting functions
library(tidyverse)
library(ggpubr)
library(ggExtra)
```

## Load and process data
Data can be downloaded from ArrayExpress (Accession Number: ).

```{r}
input_dir <- "/Volumes/LACIE/RPE_scRNA/PUBLISHED/H9_RPE_YOUNG/outs/filtered_feature_bc_matrix/"
data <- Read10X(input_dir)
seurat_obj <- CreateSeuratObject(data, project = "RPE_scRNA")

# Add metadata
seurat_obj[["condition"]] <- "Young"

seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")
seurat_obj[["percent.rb"]] <- PercentageFeatureSet(seurat_obj, pattern = "^RPS|^RPL")

```

## Extract metadata
We will use the QC metrics Seurat has calculated for us on object generation.

```{r printMeta}
# Retrieve values to plot from the Seurat object
metadata <- seurat_obj@meta.data
print(metadata[1:10,])

```

## Quality control
This function will calculate mean absolute deviation thresholds (MAD) - cells 
beyond this point are usually outliers. This function utilises the default 
`stats` R package.


```{r madFunction}
calcThreshold <- function(x,
                          nmads = 3,
                          type = c("both", "lower", "upper"),
                          na.rm = FALSE){
    med_val <- stats::median(x, na.rm = na.rm)
    mad_val <- stats::mad(x, center = med_val, na.rm = na.rm)
    upper_limit <- med_val + nmads * mad_val
    lower_limit <- med_val - nmads * mad_val
    
    if (type == "lower"){
      return(lower_limit)
    } else if (type == "upper") {
      return(upper_limit)
    } else(
      return(list(lower_limit, upper_limit))
    )
}
```

We will use this function to generate thresholds for the number of UMIs per cell.
We have to convert the counts to log10 for this to work. We can then convert this
back afterwards.

```{r calcThreshold}
# Use function to calculate 3 MAD value of total UMI
nCount_lower_threshold <- calcThreshold(log10(metadata$nCount_RNA), 
                                        nmads = 3, type = "lower")
nCount_upper_threshold <- calcThreshold(log10(metadata$nCount_RNA), 
                                        nmads = 3, type = "upper")

# Use function to calculate 3 MAD value of features
nFeature_lower_threshold <- calcThreshold(log10(metadata$nFeature_RNA), 
                                        nmads = 3, type = "lower")
nFeature_upper_threshold <- calcThreshold(log10(metadata$nFeature_RNA), 
                                        nmads = 3, type = "upper")

# Convert to original value (unlog scale)
nCount_lower_threshold <- 10^(nCount_lower_threshold)
nCount_upper_threshold <- 10^(nCount_upper_threshold)
nFeature_lower_threshold <- 10^(nFeature_lower_threshold)
nFeature_upper_threshold <- 10^(nFeature_upper_threshold)

```

We can now review the cutoffs on our plots, in addition to the percentage of 
mitochondrial and ribosomal-associated transcription.

```{r generatePlots}
# Generate plots
ncount_hist <- ggplot(metadata, aes(nCount_RNA)) + geom_histogram(bins = 100) +
  theme_bw() + geom_vline(xintercept = nCount_lower_threshold, color = "red") +
  geom_vline(xintercept = nCount_upper_threshold, color = "red") +
  ggtitle("H9 RPE (Young)", subtitle = "Total UMIs per cell") +
  xlab("Total UMIs per cell") + ylab("Number of cells")

nfeature_hist <- ggplot(metadata, aes(nFeature_RNA)) + 
  geom_histogram(bins = 100) +
  theme_bw() + geom_vline(xintercept = nFeature_lower_threshold, color = "red") + 
  geom_vline(xintercept = nFeature_upper_threshold, color = "red") + 
  ggtitle("H9 RPE (Young)", subtitle = "Total genes per cell") + 
  xlab("Total genes per cell") + ylab("Number of cells")

mt_hist <- ggplot(metadata, aes(percent.mt)) + geom_histogram(bins = 100) +
  theme_bw() + geom_vline(xintercept = 20, color = "red") + 
  ggtitle("H9 RPE (Young)", subtitle = "Mitochondrial expression") +
  xlab("% Mitochondrial expression") + ylab("Number of cells")

rb_hist <- ggplot(metadata, aes(percent.rb)) + geom_histogram(bins = 100) +
  theme_bw() + geom_vline(xintercept = 60, color = "red") + 
  ggtitle("H9 RPE (Young)", subtitle = "Ribosomal expression") +
  xlab("% Ribosomal expression") + ylab("Number of cells")

ggarrange(ncount_hist, nfeature_hist, mt_hist, rb_hist, ncol = 2, nrow = 2)
```

We also need to review all these metrics in the context of each other. 

```{r, qcontext}
count_feature_plot <- ggscatter(metadata, x = "nCount_RNA", 
                                y = "nFeature_RNA", 
                                color = "percent.mt", size = 1, 
                                title = "H9 RPE (Young) QC Plot") + 
  geom_hline(yintercept = nFeature_lower_threshold, color = "#94221f") + 
  geom_hline(yintercept = nFeature_upper_threshold, color = "#94221f") + 
  geom_vline(xintercept = nCount_lower_threshold, color = "#94221f") + 
  geom_vline(xintercept = nCount_upper_threshold, color = "#94221f") +
  gradient_color(c("#1565C0", "#b92b27"))
count_feature_plot <- ggpar(count_feature_plot, 
                            legend = "bottom", legend.title = "% Mt Expression")
count_feature_plot <- ggMarginal(count_feature_plot, type = "histogram")
count_feature_plot
```

The upper limit threshold for features may be too low, as it will exclude a
number of cells that are a part of the main body of cells. We will set the 
feature threshold manually to 5000. The fraction of mitochondrial expression
metric shows most of the high expressors are excluded by the total UMI and
detected feature lower thresholds. Togther they show these cells are likely
dead or dying cells.

```{r filterCells}
# Original number of cells
ncells1 <- ncol(seurat_obj)

# Filter poor quality cells from dataset
seurat_obj <- subset(seurat_obj, nCount_RNA > nCount_lower_threshold &
                       nCount_RNA < nCount_upper_threshold &
                       nFeature_RNA > 220 &
                       nFeature_RNA < 5000 &
                       percent.mt < 25 &
                       percent.rb < 60)

ncells2 <- ncol(seurat_obj)
ncelldiff <- ncells1 - ncells2 
print(sprintf("%d cells filtered from dataset.", ncelldiff))
```

## Normalisation

```{r sctransform, results="hide", error=FALSE, message=FALSE, warning=FALSE}
seurat_obj <- SCTransform(seurat_obj, 
                          vars.to.regress = c("percent.mt", "percent.rb"),
                          conserve.memory = FALSE, verbose = FALSE)
```

## Dimensionality reduction
```{r dimreduction}
# Principal component analysis
seurat_obj <- RunPCA(seurat_obj, verbose = FALSE)

# UMAP
seurat_obj <- RunUMAP(seurat_obj, dims = 1:30, umap.method = "uwot")

```

```{r dim_plots, fig.width=6}
DimPlot(seurat_obj, reduction = "pca")
DimPlot(seurat_obj, reduction = "umap")
```

## Clustering

```{r clustering}
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:30)
windows <- c(seq(0, 1, by = 0.1))

for (res in windows){
  seurat_obj <- FindClusters(seurat_obj, resolution = res, verbose = FALSE)
}
```

Build a clustree to review cluster results at different resolutions. 

```{r clustree, fig.width = 6, fig.height = 6}
library(clustree)
clustree_plot <- clustree(seurat_obj, prefix = "SCT_snn_res.")
clustree_plot
```

The left side of the tree show the cluster identities are unstable, while the 
right branch is very stable. Stability of the branches are represented by the
arrows moving from node to node. The goal is to minimise the number of 
unclassified cells that are in cluster 0.

```{r check_res, fig.width = 6}
metadata <- FetchData(seurat_obj, vars = c("seurat_clusters", "SCT_snn_res.0.6"))
res_counts <- as.data.frame(table(metadata$SCT_snn_res.0.6))
res_hist <- ggplot(res_counts, aes(x = Var1, y = Freq)) + 
  geom_bar(stat = "identity") + theme_bw() + 
  ggtitle("Number of cells per cluster", subtitle = "Cluster Resolution: 0.6")
res_hist
```

