---
title: "Trajectory analysis of H9 RPE samples"
output:
  html_document:
    df_print: paged
---

## Monocle3 Analysis of integrated H9 RPE samples
### R Environment Setup
```{r}
library(Seurat)
library(monocle3)
library(SingleCellExperiment)
library(tidyverse)
library(ggpubr)
```

### Loading input files
```{r extractMonocle3}
# Load integrated object
seurat_obj <- readRDS("/Volumes/LACIE/RPE_scRNA/PUBLISHED/H9_RPE_IntegratedObject.rds")

# Load colours for clusters
colour_df <- read_csv("~/Documents/Experiments/RPE_scRNA/Analysis/RevisedAnalysis/cluster_colours.csv")
colour_vec <- as.character(colour_df$Colour)
names(colour_vec) <- as.character(colour_df$Cluster)
```

### Building Monocle 3 from Seurat object
```{r extractData}
# Retrieve residuals from integrated slot
counts <- seurat_obj[["integrated"]]@scale.data

# Retrieve metadata 
metadata <- FetchData(seurat_obj, c("condition", "Months", "cluster", 
                                    "seurat_clusters",
                                     "nCount_RNA", 
                                    "nFeature_RNA", 
                                    "percent.mt", 
                                    "percent.rb"))
colnames(metadata)[3] <- "cluster_identity"

# Create gene dataframe
gene_annotation <- data.frame(gene_short_name = rownames(counts), 
                              num_cells_expressed = 
                                Matrix::rowSums(counts > 0), 
                              row.names = rownames(counts))

# Create CDS monocle object
cds <- new_cell_data_set(counts,
                         cell_metadata = metadata,
                         gene_metadata = gene_annotation)
```

As this is scRNA-seq, we will use PCA to reduce the dimensionality of the data.
We will use 30 dimensions, as used in previous analyses. Note that as we are 
using residuals, normalization and scaling is not necessary.

```{r dimreduction, fig.width = 7}
cds <- preprocess_cds(cds, num_dim = 30, norm_method = "none", scaling = FALSE)
plot_pc_variance_explained(cds)
```

We are bypassing the `align_cds` step as we are using harmonized Pearson 
residuals from the Seurat workflow. The PCA-reduced data is reduced further
for plotting, using UMAP.

```{r run_umap, fig.width = 7}
cds <- reduce_dimension(cds, max_components = 2, reduction_method = "UMAP",
                        cores = 1, preprocess_method = "PCA", verbose = FALSE)
plot_cells(cds, reduction_method = "UMAP", 
           color_cells_by = "Months", 
           show_trajectory_graph = FALSE) +
          scale_color_manual(name = "Months", values = c("1" = "#e71837",
                                                 "12" = "#00674f")) +
          ggtitle("Timepoint", subtitle = "Monocle 3 UMAP")
```

Perform clustering and learn the trajectory on the resulting UMAP projection.

```{r cluster_cells, fig.width = 7}
cds <- cluster_cells(cds)
cds <- learn_graph(cds)

cluster_monocle <- plot_cells(cds, reduction_method = "UMAP", 
           color_cells_by = "cluster", 
           show_trajectory_graph = TRUE) +
  ggtitle("Monocle 3 Clusters", subtitle = "Monocle 3 Trajectory Analysis")


partition_monocle <- plot_cells(cds, reduction = "UMAP",
                             color_cells_by = "partition",
                             show_trajectory_graph = TRUE) +
  ggtitle("Partitions", subtitle = "Monocle 3 Trajectory Analysis")

cluster_seurat <- plot_cells(cds, reduction = "UMAP",
                             color_cells_by = "cluster_identity",
                             show_trajectory_graph = TRUE) +
    scale_colour_manual(name = "Cluster", values = colour_vec) +
  ggtitle("Clusters", subtitle = "Monocle 3 Trajectory Analysis")

ggarrange(cluster_monocle, partition_monocle, cluster_seurat, ncol = 3)
```

The analysis shows a branching trajectory.

From previous analyses, we know proliferative cells are still present in the
population. We can use these as a starting point for the trajectory.

```{r ki67, fig.width = 7}
proliferative_genes <- c("MKI67", "TOP2A", "TPX2", "PTTG1", "PCLAF", "RRM2")
proliferative_monocle_umap <- plot_cells(cds, genes = proliferative_genes,
                                         show_trajectory_graph= TRUE,
                                         label_branch_points = FALSE, 
                                         label_leaves = FALSE,
                                         label_roots = FALSE,
                                         label_cell_groups=FALSE,
                                         label_groups_by_cluster = FALSE)
proliferative_monocle_umap
```

To calculate pseudotime values, we will set the root of the tree at the point
closest to the proliferative cells. THis is done via a shiny app, so we will not
run this chunk.

```{r calcPseudotime, eval = FALSE}
# Use shiny app to select node closest to proliferative cells
cds <- order_cells(cds)
```

We can then review the pseudotime values.
```{r show_pseudotime}
cds <- readRDS("/Volumes/LACIE/RPE_scRNA/PUBLISHED/H9_RPE_Residual_Monocle3_Object.rds")

# Plot pseudotime values over UMAP
pseudotime_plot <- plot_cells(cds, color_cells_by = "pseudotime",
                           show_trajectory_graph = TRUE,
                           label_cell_groups = FALSE,
                           label_leaves = FALSE,
                           label_branch_points = FALSE,
                           label_roots = FALSE) +
  ggtitle("Clusters", subtitle = "Monocle 3 Trajectory Analysis")

pseudotime_plot
```

To identify genes that change as a function of pseudotime, we will perform
the Moran I's test via the `graph_test` function. 

```{r graph_test_show, eval = FALSE}
trajectory_pr_test_df <- graph_test(cds, 
                                   neighbor_graph="principal_graph", 
                                   cores=4)
trjaectory_pr_test_df <- trajectory_pr_test_df %>% 
  rownames_to_column(var = "gene_id")
trajectory_pr_de_genes_df <- trjaectory_pr_test_df %>% filter(q_value < 0.05)
trajectory_pr_de_genes_df <- trajectory_pr_de_genes_df %>% 
  arrange(desc(abs(morans_I))) %>%
  dplyr::select(gene_id, gene_short_name, status, everything())
```

```{r graph_test, include = FALSE}
trajectory_pr_test_df <- graph_test(cds, 
                                   neighbor_graph="principal_graph", 
                                   cores=4)
trjaectory_pr_test_df <- trajectory_pr_test_df %>% 
  rownames_to_column(var = "gene_id")
trajectory_pr_de_genes_df <- trjaectory_pr_test_df %>% filter(q_value < 0.05)
trajectory_pr_de_genes_df <- trajectory_pr_de_genes_df %>% 
  arrange(desc(abs(morans_I))) %>%
  dplyr::select(gene_id, gene_short_name, status, everything())
```

```{r print_results}
trajectory_pr_de_genes_df
```

Using significant genes (FDR < 0.05), we identified co-expression modules at 
multiple resolutions. The function will select the most stable result.

```{r coexpression_modules}
gene_module_df <- find_gene_modules(cds[trajectory_pr_de_genes_df$gene_id,], 
                                    reduction_method = "UMAP", 
                                    random_seed = 1, cores = 3, 
                                    resolution = 10^seq(-6, -1))
gene_module_df <- gene_module_df %>% arrange(supermodule, module)
gene_module_df
```

We will review how these modules relate to each cluster by generating a heatmap.

```{r heatmap, fig.width = 7}
cell_group_df <- tibble::tibble(cell = rownames(colData(cds)),
                                cell_group = colData(cds)$cluster_identity)

agg_mat <- aggregate_gene_expression(cds, gene_module_df, cell_group_df)
row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))

agg_mat <- as.matrix(agg_mat)
agg_mat <- scale(agg_mat)

cols = rev(colorRampPalette(RColorBrewer::brewer.pal(11,"RdYlBu"))(100))
breaks = seq(0, 1, length=101)

library(ComplexHeatmap)
heatmap <- Heatmap(agg_mat, 
                   cluster_rows = function(m) hclust(dist(agg_mat), method = "ward.D2"),
                   row_names_gp = gpar(fontsize = 8),
                   col = cols)
heatmap
```
